---
title: "Discussion 4"
author: "Nadine Ramirez"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

**Question 2: Singular Value Decomposition (SVD)**

```{r echo=FALSE}
library(readr)
SP <- read_csv("StudentsPerformance.csv")
View(SP)
head(SP, 10)

```
I chose the dataset on student performances which detail the relationship between students' backgrounds and performances (on test scores). 

```{r, echo = FALSE}
#The following code will be converting categories into num values 
SP$lunch_num <- as.numeric(factor(SP$lunch, levels = c("free/reduced", "standard")))
SP$gender_num <- as.numeric(factor(SP$gender, levels = c("female", "male"))) 
SP$race.ethnicity_num <- as.numeric(factor(SP$race.ethnicity, levels = c("group A", "group B", "group C", "group D", "group E"))) 
SP$parental.level.of.education_num <- as.numeric(factor(SP$parental.level.of.education, levels = c("some college", "high school", "associate's degree", "master's degree", "bachelor's degree", "some high school")))
SP$test.prep.course_num <- as.numeric(factor(SP$test.prep.course, levels = c("completed", "none")))

```


```{r}
#Only keeping the numerical data
SP<-subset(SP,select=-c(gender,race.ethnicity,lunch,parental.level.of.education,test.prep.course))
head(SP,10)
```

Singular Value Decomposition
```{r}
sv = svd(SP)
sv$d
```

**Histogram**
The histogram shows the total score distribution of the sample of students. This lead for the range of 50 - 250 to be created for the reduced
```{r}
library(ggplot2)
hist(SP$total.score)
```





```{r}
library(ggplot2)
library(plotly)
```


```{r}
reducedSP = SP[which( SP$total.score >= 50 & SP$total.score <= 300),]
dim(reducedSP)

plot_ly(reducedSP, x = ~math.score, y = ~writing.score, z = ~reading.score,size = 1) %>% add_markers(color = ~total.score)
```

**Single Value Reduced**
```{r}
svd(reducedSP)$d

```
**Centered Reduced**
```{r}
centerreducedSP=scale(reducedSP, center=TRUE,scale=FALSE)
cov(centerreducedSP)

```
**Maximum:**
```{r}
max(abs(centerreducedSP))
```
```{r}
usv= svd(centerreducedSP)
```

Approximations:
*First Approximation*
```{r}
approximationSP1=usv$d[1] * usv$u[, 1]%*%t(usv$v[, 1])
max(abs(centerreducedSP- approximationSP1))
```
*Second Approximation*

```{r}
approximationSP2= usv$d[1] * usv$u[, 1]%*%t(usv$v[, 1]) + usv$d[2]*usv$u[, 2]%*%t(usv$v[, 2]) 
max(abs(centerreducedSP- approximationSP2))
```

*Third Approximation*
```{r}
approximationSP3= usv$d[1] * usv$u[, 1]%*%t(usv$v[, 1]) + usv$d[2]*usv$u[, 2]%*%t(usv$v[, 2]) +usv$d[3] * usv$u[, 3]%*%t(usv$v[, 3])
max(abs(centerreducedSP- approximationSP3))
```

**Plot**

```{r}
plotSP= as.data.frame(cbind(usv$d[1] * usv$u[, 1],usv$d[2] * usv$u[, 2], usv$d[3] * usv$u[, 3]))
plot_ly(plotSP,x= ~plotSP[,1], y= ~plotSP[,2], z=~plotSP[,3], size = 1) %>% add_markers(color = ~reducedSP$total.score)
```

This leaves us with a large variance which can be seen by the shape of the graph (no true shape)

**Centered and Scaled**
```{r}
centerscaleSP=scale(reducedSP, center=TRUE,scale=TRUE)
cov(centerscaleSP)
```

**Final Centered Plot**
```{r}
usv= svd(centerscaleSP)
p2SP= as.data.frame(cbind(usv$d[1] * usv$u[, 1],usv$d[2] * usv$u[, 2], usv$d[3] * usv$u[, 3]))
plot_ly(p2SP,x= ~p2SP[,1], y= ~p2SP[,2], z=~p2SP[,3], size = 1) %>% add_markers(color = ~reducedSP$total.score)
```

